Exploring dimensions of biodiversity in Japanese ferns

### 2021-08-02 ms/SI.Rmd
Update MS
add bioregions dendrogram, phylo. tree

# Load packages only used when knitting Rmd
library(kableExtra)
library(scales)
```

@@

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp",
  "morans_i",
  "t_test_results",
  "aic_env_repro",
  "comm_ferns",
  "repro_data"
))
```

@@

@@

>>>>

# Load packages only used when knitting Rmd
library(kableExtra)
library(scales)
library(ggtree)
```

@@

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp",
  "morans_i",
  "t_test_results",
  "aic_env_repro",
  "comm_ferns",
  "repro_data",
  "japan_pterido_tree"
))
```

@@

```{r tree, fig.width = 8, fig.height = 8}
phy <- japan_pterido_tree

phy$tip.label <- str_replace_all(phy$tip.label, "_", " ")

# Make a dataframe (tibble) with node IDs (integers) and their corresponding bootstrap support values
# The tibble has two columns: one called "node", the other can be named as we like (here, "bootstrap")
bs_tibble <- tibble(
  # hard-code node ID: internal nodes start after tip nodes,
  # and phy$node.label is in the same order as internal nodes
  node=1:Nnode(phy) + Ntip(phy), 
  # Don't print BS < 50%
  # (or to show all BS, just do `bootstrap = phy$node.label`)
  bootstrap_raw = phy$node.label,
  bootstrap = parse_number(bootstrap_raw))

p <- ggtree(phy, layout = "fan", size = 0.25, open.angle = 3) %<+% 
  bs_tibble +
  geom_nodepoint(aes(color = bootstrap), size = 0.2) +
  geom_tiplab(size = 0.6)  +
  theme(legend.justification = c(1, 0), legend.position = c(1, 0)) +
  scale_color_scico(palette = "turku", direction = -1, na.value = "grey", name = "BS") +
  geom_treescale(x = 220, y = 708, width = 100, fontsize = 0)

rotate_tree(p, 3)
```

**`r s_figure("tree")`**. Time tree of native Japanese ferns and lycophytes, excluding hybrids. Points at nodes indicate bootstrap support. Scalebar 100 my.

@@

```{r dendrogram-a, fig.width = 7, fig.height = 9}
# Extract dendrogram from phyloregions analysis, convert to phylogeny
dendro_tax <- attributes(k_taxonomy$df)$meta$hclust.obj %>% as.phylo() %>% ladderize()

# Map taxonomic clusters to grid cells
clusters_tax <-
  biodiv_ferns_spatial %>%
  as_tibble() %>%
  select(grids, cluster = taxonomic_cluster) %>%
  # Convert cluster to numeric to use max() later
  mutate(cluster = levels(cluster)[cluster] %>% as.numeric)

cluster_nodes_tax_tbl <- map_df(1:max(clusters_tax$cluster), ~map_cluster_to_nodes(dendro_tax, clusters_tax, .)) %>%
  mutate(cluster = as.factor(cluster))

# Use the ggtree::`%<+%` operator to map the bootstrap values onto the ggtree object
ggtree(dendro_tax) %<+% 
  cluster_nodes_tax_tbl +
  aes(color = cluster) +
  scale_color_manual(
    values = bioregion_cols, na.value = "grey", name = "Cluster") +
  theme(legend.position = "right") +
  labs(subtitle = "**A** Taxonomic") +
  theme(plot.subtitle = element_markdown())
```

**`r s_figure("dendrogram")`**. Bioregion dendrogram. Tips are sites (names not shown); color shows cluster; internal branches between clusters 'NA'. **A**, taxonomic bioregions. **B**, phylogenetic bioregions. 
\clearpage

```{r dendrogram-b, fig.width = 7, fig.height = 9}
# Extract dendrogram from phyloregions analysis, convert to phylogeny
dendro_phy <- attributes(k_phylogeny$df)$meta$hclust.obj %>% as.phylo() %>% ladderize()

# Map taxonomic clusters to grid cells
clusters_phy <-
  biodiv_ferns_spatial %>%
  as_tibble() %>%
  select(grids, cluster = phylo_cluster) %>%
  # Convert cluster to numeric to use max() later
  mutate(cluster = levels(cluster)[cluster] %>% as.numeric)

cluster_nodes_phy_tbl <- map_df(1:max(clusters_phy$cluster), ~map_cluster_to_nodes(dendro_phy, clusters_phy, .)) %>%
  mutate(cluster = as.factor(cluster))

# Use the ggtree::`%<+%` operator to map the bootstrap values onto the ggtree object
ggtree(dendro_phy) %<+% 
  cluster_nodes_phy_tbl +
  aes(color = cluster) +
  scale_color_manual(values = bioregion_cols, na.value = "grey", name = "Cluster") +
  theme(legend.position = "right") +
  labs(subtitle = "**B** Phylogenetic") +
  theme(plot.subtitle = element_markdown())
```

**`r s_figure("dendrogram")`**, continued.

### 2021-08-02 ms/manuscript.Rmd
Update MS
add bioregions dendrogram, phylo. tree

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent`) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families of native Japanese ferns. 
The topology was in general agreement with other recent, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a]. 
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* (families follow @PteridophytePhylogenyGroupI2016). 
The age of the fern clade is estimated to be `r phylogeny_age` ma.

@@

Analysis of bioregions grouped grid-cells into `r english2(k_taxonomy$optimal$k)` or `r english2(k_phylogeny$optimal$k)` clusters (`r s_figure("k-plot")`) on the basis of taxonomic or phylogenetic distances, respectively. 
The vast majority of grid-cells fell to a subset of these clusters: using taxonomic distances, they include Cluster 1 (Hokkaido and northern Honshu), Cluster 2 (southern Honshu), Cluster 3 (Ryukyu Islands, excluding Yakushima), and Cluster 4 (Ogasawara Islands) (`r figure("bioregions")`A). 
Phylogenetic distances produced similar results, but the Ryukyu and Ogasawara Islands merged into a single cluster (`r figure("bioregions")`B). 
The other much smaller clusters are likely artifacts of insufficiently low taxon richness for robust clustering, and are not discussed further.

>>>>

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent`) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families of native Japanese ferns. 
The topology was in general agreement with other recent, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a] (`r s_figure("tree")`). 
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* (families follow @PteridophytePhylogenyGroupI2016). 
The age of the fern clade is estimated to be `r phylogeny_age` ma.

@@

Analysis of bioregions grouped grid-cells into `r english2(k_taxonomy$optimal$k)` or `r english2(k_phylogeny$optimal$k)` clusters (`r s_figure("k-plot")`) on the basis of taxonomic or phylogenetic distances, respectively. 
The vast majority of grid-cells fell to a subset of these clusters: using taxonomic distances, they include Cluster 1 (Hokkaido and northern Honshu), Cluster 2 (southern Honshu), Cluster 3 (Ryukyu Islands, excluding Yakushima), and Cluster 4 (Ogasawara Islands) (`r figure("bioregions")`A; `r s_figure("dendrogram")`A). 
Phylogenetic distances produced similar results, but the Ryukyu and Ogasawara Islands merged into a single cluster (`r figure("bioregions")`B; `r s_figure("dendrogram")`B). 
The other much smaller clusters are likely artifacts of insufficiently low taxon richness for robust clustering, and are not discussed further.
